Formalizing the Edmonds-Karp Algorithm

### 2016-03-07 paper/root.tex
Added 'lessons learned' section. Minor tuning.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation. The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

@@

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
Moreover, there are two distinct vertices $s,t\in V$ which are called \emph{source} and \emph{sink}.

@@

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

@@

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}%
Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
us to define named contexts that fix some parameters and assumptions.
For example, the graph theory is developed in the locale \isai{Graph}, which 
fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
\begin{lstlisting}
  locale Graph = fixes c :: "edge => 'capacity" begin
    definition "E == {(u, v). c (u, v) ~= 0}"
    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
    [...]
\end{lstlisting}
Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.

Networks are based on graphs, and add the source and sink node, as well as the network assumptions:
\begin{lstlisting}
  locale Network = Graph + fixes s t :: node
    assumes no_incoming_s: "\<forall>u. (u, s) \<notin> E"
    [...]
\end{lstlisting}
Most theorems presented in this paper are in the context of the \isai{Network} locale.

@@

As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
``Equation~26.4'' by ``definition of $\augment$'':

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}

In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

@@

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_correct}

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}, the bind operation fails.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, \isai{m<=m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
also possible results of \isai{m'}. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

@@

The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.

@@

Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.

@@

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path is no longer augmenting.

@@

  \subsection{Using Breadth First Search}
  A standard way to find a shortest path in a graph is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the Refinement Framework. Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path. For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm. Note that the resulting BFS algorithm is independent, and can be reused for shortest path finding in other applications.

@@

  \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
  In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
  in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
  Obviously, computing the complete residual graph each time before searching for an augmenting path is a bad idea. 
  One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
  the BFS, these operations are time critical.

  After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of factor two.
  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
  operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
  relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$ then yields the desired algorithm.

@@

  A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
  As an edge in the residual graph is either parallel or reverse to 
  a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
  It is straightforward to show that this implementation actually returns the successors in the residual graph.

  \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
  In a final step, we have to choose efficient data structures for the algorithm. The objects that the algorithm works with are
  capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.

@@

  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for data refinements from 
  purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
  The formalization of this refinement step consists of setting up the mappings between the abstract and concrete data structures,
  and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
  Finally, Imperative/HOL comes with a setup for the Isabelle code generator~\cite{Haft09,HaNi10} to generate imperative programs in OCaml, SML, Scala, and Haskell.

@@

  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
  for which we can export code, and have proved the following theorem: 
  \Snippet{edmonds_karp_correct}
  Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
  If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
  Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c==ln_\<alpha> el}).
  If the algorithm returns some number \isai{N} and residual graph \isai{cf}, then the input was a well-formed edge list that describes a valid network with at most \isai{N} nodes.
  Moreover, the returned residual graph describes a flow \isai{f} in the network, which is maximal. 
  As the case distinction is exhaustive, this theorem states the correctness of the algorithm. 
  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed. However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
  computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations like $f~x = f~x$, 
  which is provable by reflexivity, but causes the generated program to diverge.

@@

  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, our implementation is faster by a factor of $1.2$. Note that the Java implementation operates on flows, while our implementation 
  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).

  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
  a reasonably fast verified implementation.

@@

\section{Conclusion}\label{sec:concl}
  We have presented a verification of the Edmonds-Karp algorithm, using a stepwise refinement approach.
  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, 
  specialized it to the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number of outer loop iterations.
  We then conducted several refinement steps to derive an efficiently executable implementation of the algorithm, 
  including a verified breadth first search algorithm to obtain shortest augmenting paths. 
  Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
  Our implementation compares well to a reference implementation in Java.

  Our formalization has combined several techniques to achieve an elegant and accessible formalization: 
  Using the Isar proof language~\cite{Wenzel99}, we were able to provide a completely rigorous but 
  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16}
  allowed us to present the Ford-Fulkerson method on a level 
  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
  implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
  main algorithm. The BFS algorithm can be reused as building block for other algorithms. The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of data structures 
  supported by the Sepref tool, such that it can be re-used for future formalizations.

  We conclude with some statistics:
  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm required 2600 lines.
  The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
  The remaining lines are contributed by the network checker and some auxiliary theories. The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.

  \subsection{Related Work}\label{sec:related_work}
  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations''. Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
  As far as we could understand the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
  Termination is only proved for integer valued capacities.

  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms and their implementations, using different techniques and proof assistants. Noschinski~\cite{Nosch15} verifies a checker for (non-)planarity certificates using a bottom-up approach. Starting at a C implementation,
  the AutoCorres tool~\cite{Greenaway15,GAK12} generates a monadic representation of the program in Isabelle. Further abstractions are applied
  to this representation to hide low-level details like pointer manipulations and fixed size integers. Finally, a verification condition
  generator is used to prove the abstracted program correct. Note that their approach takes the opposite direction than ours: While they start at a concrete version of the algorithm and use abstraction steps to eliminate implementation details, we start at an abstract version, and use concretization steps to introduce implementation details.

>>>>

In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation. The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

@@

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
Moreover, there are two distinct vertices $s,t\in V$, which are called \emph{source} and \emph{sink}.

@@

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) the source only has outgoing edges while the sink only has incoming edges; 2) if the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

@@

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
It then pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}%
Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
us to define named contexts that fix some parameters and assumptions.
For example, the graph theory is developed in the locale \isai{Graph}, which 
fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
\begin{lstlisting}
  locale Graph = fixes c :: "edge => capacity" begin
    definition "E == {(u, v). c (u, v) ~= 0}"
    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
    [...]
\end{lstlisting}
Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.

Networks are based on graphs, and add the source and sink nodes, as well as the network assumptions:
\begin{lstlisting}
  locale Network = Graph + fixes s t :: node
    assumes no_incoming_s: "\<forall>u. (u, s) \<notin> E"
    [...]
\end{lstlisting}
Most theorems presented in this paper are in the context of the \isai{Network} locale.

@@

As an example consider the proof that for a flow $f$ and a residual flow $f'$, the augmented flow $f\augment f'$ is again a valid flow. In particular, one has to show 
that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
``Equation~26.4'' by ``definition of $\augment$'':

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}
% 
In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

@@

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Isabelle Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_correct}

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
The bind operation fails iff either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m},

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, \isai{m <= m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
also possible results of \isai{m'}. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

@@

The Isabelle Refinement Framework also supports data refinement. The representation of results can be changed according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the set of concrete results that are related to an 
abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.

@@

Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
them down to statements that do not contain monad constructs any more. In many cases, these verification conditions reflect the core idea of the proof precisely.

@@

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path does no longer exist.

@@

  \subsection{Using Breadth First Search}
  A standard way to find a shortest path in a graph is breadth first search (BFS). Luckily, we had already formalized a BFS algorithm as an example for the 
  Isabelle Refinement Framework. Unfortunately, this algorithm only computed the minimum distance between two nodes, without returning an actual path. For this project, we extended the formalization accordingly, and added an efficient imperative implementation, using the same stepwise refinement techniques as for the main algorithm. Note that the resulting BFS algorithm is independent, and can be reused for finding shortest paths in other applications.

@@

  \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
  In a next step, we observe that the algorithm is phrased in terms of a flow, which is updated until it is maximal. 
  In each iteration, the augmenting path is searched on the residual graph induced by the current flow.
  Obviously, computing the complete residual graph in each iteration is a bad idea. 
  One solution to this problem is to compute the edges of the residual graph on the fly from the network and the current flow.
  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
  the BFS, these operations are time critical.

  After our profiling indicated a hot spot on accessing the capacity matrices of the network and the flow, we switched to 
  an algorithm that operates on a representation of the residual graph directly. This resulted in a speed-up of roughly factor two.
  As the residual graph uniquely determines the flow (and vice versa), it is straightforward to phrase the
  operations directly on the residual graph. Performing a data refinement of the flow \wrt the refinement 
  relation $\{(c_f,f) \mid \text{$f$ is a flow}\}$ then yields the desired algorithm.

@@

  A common optimization is to pre-compute an \emph{adjacency map} from nodes to adjacent nodes in the network. 
  As an edge in the residual graph is either parallel or reverse to 
  a network edge, it is enough to iterate over the adjacent nodes in the network, and check whether they are actual successors in the residual graph.
  It is straightforward to show that this implementation actually returns the successor nodes in the residual graph.

  \subsection{Using Efficient Data Structures}\label{sec:impl_data_structures}
  In a final step, we have to choose efficient data structures for the algorithm.

  %   The objects that the algorithm works with are
%   capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.

@@

  Note that we performed the last refinement step using our Sepref tool~\cite{La15,La16}, which provides tool support for refinement from the
  purely functional programs of the Isabelle Refinement Framework into imperative programs expressed in Imperative/HOL~\cite{BKHEM08}.
  The formalization of this refinement step consists of setting up the mappings between the abstract and concrete data structures,
  and then using Sepref to synthesize the Imperative/HOL programs and their refinement proofs.
  Finally, Imperative/HOL comes with a setup for the Isabelle code generator~\cite{Haft09,HaNi10} to generate imperative programs in OCaml, SML, Scala, and Haskell.

@@

  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
  for which we can export code, and have proved the following theorem:

  \vspace{\abovedisplayskip}{\raggedright\begin{minipage}{\textwidth}
  \Snippet{edmonds_karp_correct}
  \end{minipage}}\vspace{\belowdisplayskip}

  \noindent Note that this theorem is stated as a Hoare triple, using separation logic~\cite{Rey02} assertions. There are no preconditions on the input.
  If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
  Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c == ln_\<alpha> el}).
  If the algorithm returns some number \isai{N} and residual graph \isai{cf}, then the input was a well-formed edge list that describes a valid network with at most \isai{N} nodes.
  Moreover, the returned residual graph describes a flow \isai{f} in the network, which is maximal. 
  As the case distinction is exhaustive, this theorem states the correctness of the algorithm. 
  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed. However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
  computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations like $f~x = f~x$, 
  which is provable by reflexivity, but causes the generated program to diverge.

@@

  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of $1.6$, while for dense graphs, our implementation is faster by a factor of $1.2$. Note that the Java implementation operates on flows, while our implementation 
  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).
  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
  a reasonably fast verified implementation.

@@

\section{Conclusion}\label{sec:concl}
  We have presented a verification of the Edmonds-Karp algorithm, using a stepwise refinement approach.
  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, 
  specialized it to the Edmonds-Karp algorithm, and proved the upper bound $O(VE)$ for the number of outer loop iterations.
  We then conducted several refinement steps to derive an efficiently executable implementation of the algorithm, 
  including a verified breadth first search algorithm to obtain shortest augmenting paths. 
  Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
  The runtime of our verified implementation compares well to that of an unverified reference implementation in Java.

  Our formalization has combined several techniques to achieve an elegant and accessible formalization: 
  Using the Isar proof language~\cite{Wenzel99}, we were able to provide a completely rigorous but 
  still accessible proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework~\cite{LaTu12,La12} and the Sepref tool~\cite{La15,La16}
  allowed us to present the Ford-Fulkerson method on a level 
  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
  implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
  main algorithm. The BFS algorithm can be reused as building block for other algorithms. The data structures are re-usable, too: although we had to implement the array representation of (capacity) matrices for this project, it will be added to the growing library of verified imperative data structures 
  supported by the Sepref tool, such that it can be re-used for future formalizations.

  During this project, we have learned some lessons on verified algorithm development: 
  \begin{itemize}
  \item It is important to keep the levels of abstraction strictly separated.
    For example, when implementing the capacity function with arrays, one needs to show that it is only applied to valid nodes.
    However, proving that, e.g., augmenting paths only contain valid nodes is hard at this low level. 
    Instead, one can protect the application of the capacity function by an assertion--- already on a high abstraction level where it can be easily discharged. 
    On refinement, this assertion is passed down, and ultimately available for the implementation.
    Optimally, one wraps the function together with an assertion of its precondition into a new constant, which is then refined independently.
  \item Profiling has helped a lot in identifying candidates for optimization. For example, based on profiling data, we decided to delay a 
    possible deforestation optimization on augmenting paths, and to first refine the algorithm to operate on residual graphs directly.
  \item ``Efficiency bugs'' are as easy to introduce as for unverified software. For example, out of convenience, we implemented the successor list computation by
    \emph{filter}. Profiling then indicated a hot-spot on this function. As the order of successors does not matter, we invested a bit more work to make the computation tail 
    recursive and gained a significant speed-up. Moreover, we realized only lately that we had accidentally implemented and verified
    matrices with column major ordering, which have a poor cache locality for our algorithm. Changing the order resulted in another significant speed-up.
  \end{itemize}

  We conclude with some statistics:
  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson algorithm requires 3000 lines.
  The abstract Edmonds-Karp algorithm and its complexity analysis contribute 800 lines, and its implementation (including BFS) another 1700 lines.
  The remaining lines are contributed by the network checker and some auxiliary theories. The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.

  \subsection{Related Work}\label{sec:related_work}
  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations''. Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
  As far as we understood the Mizar proof script, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
  Termination is only proved for integer valued capacities.

  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms and their implementations, using different techniques and proof assistants. Noschinski~\cite{Nosch15} verifies a checker for (non-)planarity certificates using a bottom-up approach. Starting at a C implementation,
  the AutoCorres tool~\cite{Greenaway15,GAK12} generates a monadic representation of the program in Isabelle. Further abstractions are applied
  to hide low-level details like pointer manipulations and fixed size integers. Finally, a verification condition
  generator is used to prove the abstracted program correct. Note that their approach takes the opposite direction than ours: While they start at a concrete version of the algorithm and use abstraction steps to eliminate implementation details, we start at an abstract version, and use concretization steps to introduce implementation details.
